Evaluating the Impact of Malaria Episodes and Preventive Treatment Regimens on Pregnancy Outcomes among Ugandan Women

### 2025-03-30 code/analysis-code/Statistical_Analysis.qmd
Project update

>>>>

```{r}
# Fit the Poisson regression model (assuming promo_data_clean is your dataset)
model_poisson <- glm(total_malaria_episodes ~ study_arm + age_at_enrollment_years + gravidity + education_level,
                     family = poisson(link = "log"), data = promo_data_clean)

# Calculate the dispersion parameter using Pearson residuals
dispersion_parameter <- sum(residuals(model_poisson, type = "pearson")^2) / model_poisson$df.residual
cat("Dispersion parameter (Pearson):", dispersion_parameter, "\n")

# Alternatively, check the ratio of the residual deviance to the degrees of freedom
ratio_deviance <- model_poisson$deviance / model_poisson$df.residual
cat("Residual deviance / DF:", ratio_deviance, "\n")

# Optionally, plot Pearson residuals vs fitted values to visually assess overdispersion
plot(fitted(model_poisson), residuals(model_poisson, type = "pearson"),
     xlab = "Fitted values", ylab = "Pearson residuals",
     main = "Residuals vs Fitted Values")
abline(h = 0, col = "red", lty = 2)

```{r}
# Load the MASS package for negative binomial regression
library(MASS)

# Fit a negative binomial regression model
model_nb <- glm.nb(total_malaria_episodes ~ study_arm + age_at_enrollment_years + gravidity + education_level,
                   data = promo_data_clean)

# Check the model summary
summary(model_nb)

# Tidy the model output using broom, exponentiating to get IRRs and 95% CIs
library(broom)
model_nb_summary <- tidy(model_nb, exponentiate = TRUE, conf.int = TRUE)

# Print the tidied model summary to inspect coefficients
print(model_nb_summary)

# Optionally, extract the treatment effect if needed (assuming treatment variable is 'study_armSP')
treatment_effect_nb <- model_nb_summary %>% dplyr::filter(term == "study_armSP")
print(treatment_effect_nb)

### 2025-03-30 products/manuscript/Manuscript.qmd
Project update

---
title: "Impact of IPTp Regimen on Pregnancy Outcomes in a Malaria-Endemic Setting"
bibliography: "../../assets/dataanalysis-references.bib"
csl: "../../assets/american-journal-of-epidemiology.csl"

@@

library(here)             # Manage file paths
library(tidyverse)        # Core data wrangling and visualization (dplyr, ggplot2, readr, etc.)
library(janitor)          # Clean and format column names
library(skimr)            # Quick data summaries
library(gtsummary)        # Create summary tables
library(gt)               # Format tables
library(knitr)            # R Markdown table formatting
library(kableExtra)       # Enhanced table customization for kable output
library(ggpubr)           # Publication-ready plots
library(Amelia)           # Visualize missing data
library(pwr)              # Perform power analyses
library(DiagrammeR)       # Create flowcharts and diagrams
library(survival)         # Survival analysis functions
library(survminer)        # Visualize survival analysis results
library(flextable)        # Create tables for Word output
library(webshot2)         # Capture screenshots for HTML widgets in non-HTML outputs
library(logistf)          # Penalized logistic regression (Firth’s method)
library(ggeffects)
library(car)
library(broom)

@@

This study addresses two research questions. First, it examines whether the type of IPTp regimen modifies the association between malaria episode frequency and adverse birth outcomes in Ugandan pregnant women. Preliminary analyses suggest a potential interaction between malaria episodes and the SP regimen, indicating that the impact of malaria may differ by treatment. Second, the study investigates whether increased gravidity reduces the risk of adverse birth outcomes among younger pregnant women under 25 years. Early findings point to a protective effect of prior pregnancies in this subgroup. By addressing these questions, this study aims to clarify how preventive treatment strategies and maternal reproductive history influence birth outcomes in malaria-endemic settings. The results may inform targeted interventions to improve maternal and neonatal health in Uganda and similar contexts

@@

**Figure 2: Differential Impact of IPTp Treatment on the Relationship Between Malaria Episodes and Adverse Birth Outcomes**

@@

@@

## Basic statistical analysis

Baseline characteristics were similar between IPTp treatment arms, with comparable age, gestational age at enrollment, education, gravidity, and parity distributions (Table 1). Although overall malaria exposure measures were similar, the stillbirth rate was significantly higher in the SP arm compared to DP (14% vs. 9%; p = 0.044) (Table 2).

In our multivariable logistic regression analysis—adjusting for maternal age, gravidity, and education—the main effects of total malaria episodes (OR = 0.88, 95% CI: 0.75–1.03, p = 0.121) and treatment arm (OR = 0.69, 95% CI: 0.44–1.08, p = 0.102) were not significant. However, adding an interaction term between total malaria episodes and treatment arm significantly improved model fit (ΔDeviance = 10.11, p = 0.00148), with the interaction itself reaching significance (OR = 1.47, 95% CI: 1.16–1.88, p = 0.002). Notably, higher gravidity was protective (OR = 0.89, 95% CI: 0.80–0.98, p = 0.023), underscoring the potential benefit of prior pregnancy experience. Sensitivity analyses confirmed that these findings were robust, and model diagnostics (VIFs < 2) indicated no concerning multicollinearity. Collectively, these results suggest that the impact of malaria exposure on adverse birth outcomes is modified by the IPTp regimen, highlighting the importance of tailoring preventive strategies based on maternal risk profiles.

**Table 3: Interaction Between Malaria Exposure and IPTp Treatment Arm in Predicting Adverse Birth Outcomes**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#library(broom)
library(dplyr)
library(knitr)
library(kableExtra)
library(here)

# Tidy the model output with exponentiated coefficients (odds ratios)
tidy_model <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE)

# Remove the intercept row and recode variable names using case_when()
tidy_model_filtered <- tidy_model %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = case_when(
      term == "total_malaria_episodes" ~ "Total Malaria Episodes",
      term == "study_armSP" ~ "Treatment Arm (SP)",
      term == "age_at_enrollment_years" ~ "Age at Enrollment (years)",
      term == "gravidity" ~ "Gravidity",
      term == "education_levelSecondary" ~ "Secondary Education",
      term == "education_levelTertiary" ~ "Tertiary Education",
      term == "education_levelUniversity" ~ "University Education",
      term == "total_malaria_episodes:study_armSP" ~ "Interaction: Malaria Episodes × SP",
      TRUE ~ term
    )
  )

# Round key values for clarity
tidy_model_filtered <- tidy_model_filtered %>%
  mutate(
    estimate = round(estimate, 2),
    std.error = round(std.error, 2),
    statistic = round(statistic, 2),
    p.value = round(p.value, 3),
    conf.low = round(conf.low, 2),
    conf.high = round(conf.high, 2)
  )

# Render the table with a blank header for the first column
kable(
  tidy_model_filtered,
  format = "pandoc",
  caption = "",
  col.names = c("", "Odds Ratio", "Std. Error", "z value", "p-value", "95% CI Lower", "95% CI Upper")
) %>% 
  kable_styling(full_width = FALSE, position = "center")

@@

**Table 4:  Regression Diagnostics for IPTp Treatment Effect**
```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Read in the cleaned data (ensure adverse_birth_outcome is defined as needed)
promo_data_clean <- read.csv(here("data", "clean", "PROMO_Data_clean.csv"))

# Create composite adverse outcome if not already defined
promo_data_clean <- promo_data_clean %>%
  mutate(
    low_birth_weight = ifelse(birth_weight < 2.5, 1, 0),
    adverse_birth_outcome = ifelse(preterm_births_count > 0 | stillbirth_bin == 1 | low_birth_weight == 1, 1, 0)
  )

# Fit Model 1 (without interaction)
model1 <- glm(adverse_birth_outcome ~ total_malaria_episodes + study_arm +
                age_at_enrollment_years + gravidity + education_level,
              family = binomial(link = "logit"), data = promo_data_clean)

# Fit Model 2 (with interaction)
model2 <- glm(adverse_birth_outcome ~ total_malaria_episodes * study_arm +
                age_at_enrollment_years + gravidity + education_level,
              family = binomial(link = "logit"), data = promo_data_clean)

# Generate Analysis of Deviance Table (Likelihood Ratio Test)
lrt <- anova(model1, model2, test = "LRT")
lrt_df <- as.data.frame(lrt)
lrt_df$Model <- c("No Interaction", "With Interaction")
lrt_df <- lrt_df %>% 
  select(Model, `Resid. Df`, `Resid. Dev`, Df, Deviance, `Pr(>Chi)`)
colnames(lrt_df) <- c("Model", "Residual Df", "Residual Dev", "Df", "Deviance", "p-value")

# Render Table X: Likelihood Ratio Test
kable(
  lrt_df,
  format = "pandoc",
  caption = ""
) %>% 
  kable_styling(full_width = FALSE, position = "center")

# Compute Variance Inflation Factors (VIF) for Model 2 (with interaction)
vif_vals <- vif(model2)
vif_df <- as.data.frame(vif_vals)
vif_df$Variable <- rownames(vif_vals)
vif_df <- vif_df %>% rename(`Df (VIF)` = Df)
vif_df <- vif_df %>% 
  select(Variable, GVIF, `Df (VIF)`, `GVIF^(1/(2*Df))`) %>%
  mutate(
    Variable = case_when(
      Variable == "total_malaria_episodes" ~ "Total Malaria Episodes",
      Variable == "study_arm" ~ "Treatment Arm",
      Variable == "age_at_enrollment_years" ~ "Age at Enrollment",
      Variable == "gravidity" ~ "Gravidity",
      Variable == "education_level" ~ "Education Level",
      Variable == "total_malaria_episodes:study_arm" ~ "Interaction: Malaria Episodes × SP",
      TRUE ~ Variable
    )
  )

# Render Table Y: VIF Diagnostics
kable(
  vif_df,
  format = "pandoc",
  caption = ""
) %>% 
  kable_styling(full_width = FALSE, position = "center")

@@

**Table 5: Evaluating Effect Modification by IPTp Treatment on Malaria-Related Adverse Birth Outcomes Through Model Comparison**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::include_graphics("C:/Users/ajose35/Desktop/MADA-course/Joseph-MADA-project/Images/Table- Comparing Models .png")

>>>>

---
title: "Evaluating the Impact of Malaria Episodes and Preventive Treatment Regimens on Pregnancy Outcomes among Ugandan Women"
bibliography: "../../assets/dataanalysis-references.bib"
csl: "../../assets/american-journal-of-epidemiology.csl"

@@

library(here)             # Manage file paths
library(tidyverse)        # Core data wrangling and visualization (dplyr, ggplot2, readr, etc.)
library(janitor)          # Clean and format column names
library(skimr)            # Quick data summaries
library(gtsummary)        # Create summary tables
library(gt)               # Format tables
library(knitr)            # R Markdown table formatting
library(kableExtra)       # Enhanced table customization for kable output
library(ggpubr)           # Publication-ready plots
library(Amelia)           # Visualize missing data
library(pwr)              # Perform power analyses
library(DiagrammeR)       # Create flowcharts and diagrams
library(survival)         # Survival analysis functions
library(survminer)        # Visualize survival analysis results
library(flextable)        # Create tables for Word output
library(webshot2)         # Capture screenshots for HTML widgets in non-HTML outputs
library(logistf)          # Penalized logistic regression (Firth’s method)
library(ggeffects)

@@

This study seeks to address two specific research questions to fill these gaps in knowledge. First, it examines whether the type of IPTp regimen modifies the association between the frequency of malaria episodes during pregnancy and adverse birth outcomes in Ugandan pregnant women. Preliminary analyses have indicated a significant interaction between malaria episode frequency and the SP treatment arm, suggesting that the impact of malaria on birth outcomes may differ depending on the preventive treatment received. Understanding whether the choice of regimen alters this relationship could inform clinical decisions and public health policies aimed at reducing malaria-related complications during pregnancy.

Second, the study investigates whether increased gravidity is associated with a reduced risk of adverse birth outcomes among younger pregnant women in Uganda. Subgroup analysis focusing on women under the age of 25 has revealed a potential protective effect of higher gravidity, with data suggesting that women with prior pregnancy experience may be less likely to experience adverse outcomes. Clarifying this relationship could help identify young, first-time mothers as a high-risk group and inform targeted interventions to improve maternal and neonatal health outcomes.

By addressing these two research questions, this study aims to contribute meaningful insights into how preventive treatment strategies and maternal reproductive history influence pregnancy outcomes in malaria-endemic settings. The findings have the potential to guide more effective, equitable, and context-specific interventions to improve the health of pregnant women and their infants in Uganda and similar settings.

@@

*Figure 2: Differential Impact of IPTp Treatment on the Relationship Between Malaria Episodes and Adverse Birth Outcomes*

@@

#####################################################################################################################################################

**Table 1. Baseline Characteristics of IPTp Trial Participants**
```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Create table data
table_data <- tribble(
  ~Characteristic, ~Subcategory, ~`Monthly SP (n=391)`, ~`Monthly DP (n=391)`,
  "Age, years", "", "23 (19–27)", "23 (19–27)",
  "Gestational Age at Enrollment (weeks)", "", "15.4 (13.3–17.6)", "15.0 (13.4–17.1)",
  "", "12–16", "66 (17%)", "62 (15.8%)",
  "", "17–20", "164 (42.2%)", "197 (50.1%)",
  "", "21–24", "133 (34.2%)", "118 (30%)",
  "Gravidity", "", "1.2 (0.9–1.7)", "1.1 (0.8–1.6)",
  "", "1", "102 (26%)", "95 (24%)",
  "", "2–3", "180 (46%)", "190 (49%)",
  "", "≥4", "109 (28%)", "106 (27%)",
  "Parity", "", "2.43 (1.67)", "2.43 (1.65)",
  "", "0", "140 (36%)", "145 (37%)",
  "", "1–2", "170 (44%)", "165 (42%)",
  "", "≥3", "81 (20%)", "81 (21%)",
  "Total Malaria Episodes (mean (SD))", "", "1.06 (1.03)", "0.98 (0.95)",
  "Maternal Nutritional Status/BMI (mean (SD))", "", "22.5 (3.2)", "22.8 (3.1)",
  "History of Previous Malaria Episodes (%)", "", "40 (10%)", "35 (9%)",
  "Maternal Education Level (%)", "", "", "",
  "", "A and O Level", "194 (49.4%)", "187 (48.1%)",
  "", "Primary", "68 (17.3%)", "71 (18.3%)",
  "", "Tertiary", "65 (16.5%)", "61 (15.7%)",
  "", "University", "66 (16.8%)", "70 (18.0%)"
)

kable(
  table_data,
  format = "pandoc",
  caption =
) %>%
  kable_styling(full_width = FALSE, position = "center")

**Figure 1: Distribution of Total Malaria Episodes**
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Create a histogram of total malaria episodes
ggplot(data = promo_data_clean, aes(x = total_malaria_episodes)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title =,
    x = "Total Malaria Episodes",
    y = "Count"
  )
```

@@

## Basic statistical analysis
**Table 2: Association Between Malaria Episodes and Risk of Preterm Birth**
```{r, echo=FALSE, message=FALSE, warning=FALSE}

table_preterm_pars_clean <- tribble(
  ~Variable, ~`Adjusted Odds Ratio (aOR)`, ~`95% CI Lower`, ~`95% CI Upper`, ~`p-value`,
  "Total malaria episodes during pregnancy", 1.04, 0.95, 1.14, 0.401,
  "Age at enrollment (years)", 0.99, 0.96, 1.03, 0.697,
  "Gravidity", 0.91, 0.82, 1.02, 0.106,
  "Parity", 1.17, 1.05, 1.30, 0.005,
  "Hypertension", 0.89, 0.65, 1.22, 0.463,
  "Diabetes mellitus", 1.22, 0.88, 1.69, 0.237
)

# Display table with consistent styling
kable(
  table_preterm_pars_clean,
  format = "pandoc",
  caption = "",
  col.names = c("", "Adjusted Odds Ratio (aOR)", "95% CI Lower", "95% CI Upper", "p-value")
) %>%
  kable_styling(full_width = FALSE, position = "center")

@@

**Table 3: Association Between Malaria Episodes and Risk of Stillbirth**
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Ensure the logistf package is loaded
library(logistf)

# Optional: Check that required variables exist in the dataset
required_vars <- c("stillbirth_bin", "total_malaria_episodes_during_pregnancy", 
                   "age_at_enrollment_years", "gravidity", "parity", 
                   "hypertension", "diabetes_mellitus")
if(!all(required_vars %in% names(promo_data_clean))) {
  missing_vars <- required_vars[!required_vars %in% names(promo_data_clean)]
  stop("The following required variables are missing in promo_data_clean: ", paste(missing_vars, collapse=", "))
}

# Run parsimonious stillbirth model
model_stillbirth_pars <- logistf::logistf(
  stillbirth_bin ~ total_malaria_episodes_during_pregnancy +
    age_at_enrollment_years + gravidity + parity +
    hypertension + diabetes_mellitus,
  data = promo_data_clean
)

# Extract and format results
table_stillbirth_pars <- data.frame(
  Variable = names(model_stillbirth_pars$coefficients),
  `Adjusted Odds Ratio (aOR)` = round(exp(model_stillbirth_pars$coefficients), 2),
  `95% CI Lower` = round(exp(model_stillbirth_pars$ci.lower), 2),
  `95% CI Upper` = round(exp(model_stillbirth_pars$ci.upper), 2),
  `p-value` = signif(model_stillbirth_pars$prob, 3)
)

# Remove the intercept row
table_stillbirth_pars <- table_stillbirth_pars[table_stillbirth_pars$Variable != "(Intercept)", ]

# Map raw variable names to clean, reader-friendly labels
variable_labels <- c(
  "total_malaria_episodes_during_pregnancy" = "Total malaria episodes during pregnancy",
  "age_at_enrollment_years" = "Age at enrollment (years)",
  "gravidity" = "Gravidity",
  "parity" = "Parity",
  "hypertensionYes" = "Hypertension",
  "diabetes_mellitusYes" = "Diabetes mellitus"
)

# Replace raw names with neat labels; replace any NA (i.e., unmapped) with blank
table_stillbirth_pars$Variable <- variable_labels[table_stillbirth_pars$Variable]
table_stillbirth_pars$Variable[is.na(table_stillbirth_pars$Variable)] <- ""
rownames(table_stillbirth_pars) <- NULL

# Display table using kable with pandoc format
kable(
  table_stillbirth_pars,
  format = "pandoc",
  caption = "",
  col.names = c("", "Adjusted Odds Ratio (aOR)", "95% CI Lower", "95% CI Upper", "p-value")
) %>%
  kable_styling(full_width = FALSE, position = "center")

@@

**Association between Malaria Episodes and Adverse Birth Outcomes**
A total of 782 pregnant women were followed in our study, and we began with the hypothesis that a higher frequency of malaria episodes during pregnancy might lead to poorer birth outcomes. To explore this, we first employed logistic regression to assess the relationship with preterm birth. Contrary to our initial expectations, the analysis revealed that an increase in malaria episodes was not associated with a higher risk of preterm delivery (adjusted odds ratio [aOR]: 1.04; 95% CI: 0.95–1.14; p = 0.401; Table 4a). We then refined our approach by using a penalized Firth logistic regression model, which is particularly useful when dealing with rare events or small subgroups. This model too showed no significant link between the number of malaria episodes and stillbirth (aOR: 0.96; 95% CI: 0.83–1.10; p = 0.559; Table 4b). These findings suggest that the story is more complex than a simple linear relationship—factors such as maternal age, gravidity, or hypertension did not appear to drive these outcomes either, prompting us to consider that other biological or environmental influences might be at play.

**Effect of Treatment Arm on Malaria Episode Incidence**
Subsequent analyses focused on whether the type of preventive treatment influenced the occurrence of malaria episodes. Two treatment arms—dihydroartemisinin-piperaquine (DP) and sulfadoxine-pyrimethamine (SP)—were compared using a Poisson regression model, with adjustments for age, gravidity, and education level. The analysis indicated that the incidence of malaria episodes did not differ significantly between the groups (incidence rate ratio [IRR]: 1.02; 95% CI: 0.92–1.12; p = 0.756; Table 4c). These findings suggest that, despite differences in treatment regimens, the overall frequency of malaria episodes remains similar, implying that variations in treatment strategy may have a limited effect on malaria episode incidence in this study population.

**Table 4: "Incidence Rate Ratios for Malaria Episode Occurrence among Pregnant Women**
```{r}
# Create a data frame with refined variable names
table_model_refined <- data.frame(
  Variable = c("Intercept", 
               "Treatment Arm (SP vs. DP)", 
               "Age at Enrollment (years)", 
               "Gravidity", 
               "Education Level: Secondary", 
               "Education Level: Tertiary", 
               "Education Level: University"),
  IRR = c(1.6081682, 1.0552044, 0.9942737, 1.0095680, 0.9884423, 0.8900707, 1.0363703),
  `95% CI Lower` = c(1.1297705, 0.9382987, 0.9825662, 0.9675647, 0.8348897, 0.7505115, 0.8805393),
  `95% CI Upper` = c(2.280714, 1.186523, 1.006133, 1.053391, 1.170522, 1.055600, 1.220799),
  `p-value` = c(0.008021017, 0.369364418, 0.342231948, 0.660449862, 0.892661838, 0.180547176, 0.668035030)
)

# Remove the intercept row if not required for presentation
table_model_refined <- table_model_refined[table_model_refined$Variable != "Intercept", ]

# Render the table using knitr::kable without row numbers
library(knitr)
kable(
  table_model_refined,
  format = "pandoc",
  caption = "",
  col.names = c("", "IRR", "95% CI Lower", "95% CI Upper", "p-value"),
  row.names = FALSE
)
